---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mesa

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/mesa)](https://CRAN.R-project.org/package=mesa)
[![CRAN downloads](http://cranlogs.r-pkg.org/badges/grand-total/mesa?color=blue)](https://cran.r-project.org/package=mesa)
[![R-CMD-check](https://github.com/shah-in-boots/mesa/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/shah-in-boots/mesa/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/shah-in-boots/mesa/actions/workflows/pkgdown.yaml/badge.svg)](https://github.com/shah-in-boots/mesa/actions/workflows/pkgdown.yaml)
[![test-coverage](https://github.com/shah-in-boots/mesa/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/shah-in-boots/mesa/actions/workflows/test-coverage.yaml)
[![Codecov test coverage](https://codecov.io/gh/shah-in-boots/mesa/branch/main/graph/badge.svg)](https://app.codecov.io/gh/shah-in-boots/mesa?branch=main)
[![Github commit frequency](https://img.shields.io/github/commit-activity/w/shah-in-boots/mesa)](https://github.com/shah-in-boots/mesa/graphs/commit-activity) 
<!-- badges: end -->

## Installation

The development version can be installed from [Github](https://github.com/shah-in-boots/mesa). Once `{mesa}` is accepted on CRAN, it can be installed from CRAN as well.

```{r, eval = FALSE}
# Development version
remotes::install_github("shah-in-boots/mesa")
# CRAN installation after release
install.packages("mesa")
```

## Introduction

The package `{mesa}` (Models for Epidemiology and Statistical Analysis) was intended as a way to handle causal- and epidemiology-based modeling by the following principles:

1. Role determination of variables
2. Generativity in formula creation
3. Multiple model management

## Usage

The package is simple to use. 
The `mtcars` dataset will serve as the example, and we will use linear regressions as the primary test.
This toy example shows that we will be building six models in parallel, with the key exposure being the __wt__ term, and the two outcomes being __mpg__ and __hp__.

```{r}
library(mesa)

f <- fmls(mpg + hp ~ .x(wt) + disp + cyl + am, pattern = "parallel")
m <- fit(f, .fn = lm, data = mtcars, raw = FALSE)
mt <- model_table(mileage = m)
print(mt)
```

## Classes

There are several important extended classes that this package introduces, however they are primarily used for internal validation and for shortcuts to allow more effective communication.

- `fmls` are a *version* of the base `R` formula object, but contain additional information and have extra features
- `tm` are atomic elements used to describe individual variables, and departs from how terms are generally treated in the `{stats}` package
- `mdl` and `mdl_tbl` exist primarily as *tidy* versions of class regression modeling

## Advanced Usage

The [`{mesa}`](https://cran.r-project.org/package=mesa) package is intended to be flexible, extensible, and easy-to-use (albeit opinionated). 
Please see the vignettes for additional information.
