---
title: "6. The Causal Reasoning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{6. The Causal Reasoning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mesa)
```

The other five vignettes teach the mechanics: runes, patterns, verbs, tables.
This one is the intellectual home the mechanics serve. `{mesa}` exists
because a plain `R` formula cannot say what an epidemiologist means by
`outcome ~ exposure + confounder` — it can only say that three variables sit
in some algebraic relationship. The role a variable plays is where the
*meaning* lives, and this vignette is about where that meaning comes from and
what it does and does not buy you.

## A formula is not a causal claim

`stats::formula()` knows left-hand side and right-hand side. It has no
concept of exposure, confounder, or mediator — those are epidemiological
ideas, not syntactic ones, and R has always required the analyst to carry
them in their head instead of in the object. `{mesa}`'s central move is to
make that carrying explicit: a `tm` is a variable *plus its causal role*
(`vignette("terms")`), and every layer above it — expansion, fitting,
display — reads that role rather than re-deriving intent from position.

This is deliberately modest. Attaching a role to a term does not verify that
the role is correct, any more than naming a variable `confounder` in your own
analysis script would. `{mesa}` gives the causal reasoning somewhere to live
and something to act on; the reasoning itself — is this variable really a
confounder, does this DAG actually represent the world — remains the
analyst's.

## Association is not causation, but it is not arbitrary either

Austin Bradford Hill's 1965 address to the Royal Society of Medicine
(Hill 1965) is the origin of the modern instinct that an association needs
more than a p-value before it is treated as causal. Hill offered nine
viewpoints — strength, consistency, specificity, temporality, biological
gradient, plausibility, coherence, experiment, and analogy — not as a
checklist to be scored, but as considerations that, taken together, make an
association more or less credible as a causal one.

`{mesa}`'s role vocabulary does not encode Hill's viewpoints directly — they
are arguments made *about* an association, not properties of a formula. But
two of them shape the package's structure directly:

- **Temporality** is why `strata` (`.s()`) and `random` (`.r()`) are
  deliberately *not* covariates: a variable that defines subpopulations, or
  that groups repeated observations, is not competing with the exposure for
  a coefficient — it is a fact about the data's structure that has to be
  respected before an association is even estimated.
- **Biological gradient and specificity** are why a categorical term keeps
  its levels rather than collapsing to a single coefficient
  (`vignette("mesa")`'s `"levels"` layout, and `estimate_interaction()`'s
  per-level estimates) — a dose-response pattern, or an effect that is
  specific to one subgroup, is invisible in a table that only shows "yes/no."

## Roles as a lightweight stand-in for a graph

Judea Pearl's *do-calculus* (Pearl 2010) formalizes the difference between
seeing and doing: `P(Y | X = x)`, an observed association, is not in general
equal to `P(Y | do(X = x))`, the effect of an intervention. The tool that
makes the difference legible is the causal directed acyclic graph (DAG) — a
picture of which variables cause which, from which the confounders, the
mediators, and the valid adjustment sets can be *read off* rather than
guessed at.

`{mesa}` does not draw or check a DAG (that integration is on the horizon —
see the blueprint's "DAG integration" entry). What it borrows from the
do-calculus framework is the vocabulary a DAG produces: `exposure`,
`confounder`, `mediator`, `interaction` are exactly the node-types a causal
DAG distinguishes, and a `tm`'s role is a direct, if unverified, claim about
where that term would sit in one. Declaring `.c(smoking)` is the formula-level
equivalent of drawing an arrow from `smoking` into both the exposure and the
outcome; `{mesa}` takes the claim on faith and acts on it (adjustment
displays, table layout), the same way a regression takes on faith that the
formula it is handed is the right one to fit.

## Adjustment, mediation, and the total-effect / direct-effect distinction

The most consequential place the role vocabulary shapes actual output is the
`mediator` role. Adjusting for a common cause of exposure and outcome
(a confounder) and adjusting for a variable on the causal *pathway* between
them (a mediator) are both "adding a covariate" in ordinary regression
syntax — and they answer completely different questions. Conditioning on a
mediator blocks part of the very effect under study.

VanderWeele and Robins's work on mediation (see also VanderWeele's 2015 text,
*Explanation in Causal Inference*) formalizes this as decomposing a total
effect into a **direct effect** (the exposure's effect not through the
mediator) and an **indirect effect** (the effect that travels through it). A
term wrapped `.m()` triggers exactly this decomposition in `fmls()`
(`vignette("formulas")`): instead of one adjusted formula, the family expands
into the triad

```{r}
fmls(mpg ~ .x(wt) + .m(qsec) + hp) |> formula()
```

— the total effect with the mediation pathway held open, the exposure's
effect on the mediator, and the mediator's effect on the outcome in the
exposure's presence. Comparing the exposure's coefficient across the first
and third formulas is what lets an analyst judge how much of the total
effect travels through the mediator, rather than accidentally reporting a
direct effect while believing it is a total one.

## Effect modification is a property of the exposure's effect, not a coefficient

VanderWeele and Robins (2007) distinguish four types of effect modification
based on where the modifying variable sits on the causal DAG relative to the
exposure and outcome — a distinction that matters because "the interaction
term's p-value is significant" is a much weaker claim than "the exposure's
effect differs meaningfully across levels of this modifier."

`{mesa}`'s `interaction` role (`.i()`) and `estimate_interaction()` are built
around the second framing. Rather than reporting the interaction
*coefficient* — which describes a difference in a difference, not an effect
either subgroup experiences — `estimate_interaction()` derives the exposure's
*own* effect within each level of the modifier, from the fitted coefficients'
variance-covariance structure (Figueiras, Domenech-Massons, and Cadarso 1998):
within the reference level the effect is the exposure's own coefficient;
within level *j* it is that coefficient plus the level's interaction term,
with variance `var(b_exposure) + var(b_j) + 2 cov(b_exposure, b_j)`. That is
the number a clinician or policymaker can act on — "the effect in men" and
"the effect in women" — where the raw interaction coefficient is not.

```{r}
d <- mtcars
d$cyl <- factor(d$cyl)
mt <-
  fmls(mpg ~ .x(hp) + .i(cyl)) |>
  fit(.fn = lm, data = d, raw = FALSE) |>
  model_table(data = d)
estimate_interaction(mt, exposure = "hp", interaction = "cyl")
```

## What the package does not do

It is worth being explicit about the boundary. `{mesa}` organizes causal
*intent* — it makes roles first-class, keeps mediation and effect
modification from collapsing into "just another covariate," and lays the
resulting estimates out so the distinctions survive into a published table.
It does not:

- verify that a declared confounder actually blocks a backdoor path,
- check a DAG for consistency, or
- adjudicate between competing causal assumptions.

Those remain exactly as much the analyst's responsibility as they would be
without the package. What changes is that the *decisions* — this is the
exposure, this is on the causal pathway, this is a candidate effect
modifier — are recorded once, in the formula itself, and every layer above it
(expansion, fitting, the published table) inherits them consistently instead
of silently re-deriving or contradicting them.

## References

- Hill, A. B. (1965). The Environment and Disease: Association or Causation?
  *Proceedings of the Royal Society of Medicine*, 58(5), 295–300.
  <https://doi.org/10.1177/003591576505800503>
- Pearl, J. (2010). An Introduction to Causal Inference. *The International
  Journal of Biostatistics*, 6(2), Article 7.
  <https://doi.org/10.2202/1557-4679.1203>
- VanderWeele, T. J., & Robins, J. M. (2007). Four types of effect
  modification: a classification based on directed acyclic graphs.
  *Epidemiology*, 18(5), 561–568.
  <https://doi.org/10.1097/EDE.0b013e318127181b>
- VanderWeele, T. J. (2015). *Explanation in Causal Inference: Methods for
  Mediation and Interaction*. Oxford University Press.
- Figueiras, A., Domenech-Massons, J. M., & Cadarso, C. (1998). Regression
  models: calculating the confidence interval of effects in the presence of
  interactions. *Statistics in Medicine*, 17, 2099–2105.
