---
title: "4. Fitting and Model Tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Fitting and Model Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(epigram)
```

A family of formulas is still just instructions. Layer 3 answers "what
happens when a formula meets data and a fitting approach?" (`fit()`), and
layer 4 answers "how do I store, recall, and compare many models?"
(`mdl_tbl`). This vignette covers both, because in practice they are always
used together: fit, then collect.

```{r}
d <- mtcars
d$cyl <- factor(d$cyl)
d$am <- factor(d$am, labels = c("Automatic", "Manual"))
```

## Fitting is plan-then-execute

`fit()` accepts a family, a fitting function (`.fn`), and the data. It first
builds an inspectable plan — every formula crossed with every stratum level
and every subset — with `plan_fit()`, then executes it row by row:

```{r}
f <- fmls(mpg ~ .x(wt) + hp + cyl, pattern = "sequential")
plan_fit(f, data = d)
```

`.fn` resolves **by name**, so argument order never matters, and it accepts
three things: a plain function (`lm`), a string naming one (`"lm"`), or a
`{parsnip}` model specification.

```{r, eval = requireNamespace("parsnip", quietly = TRUE)}
spec <- parsnip::linear_reg() |> parsnip::set_engine("lm")
fit(fmls(mpg ~ .x(wt) + hp), .fn = spec, data = d)
```

## Wrapped models, not raw fits

By default, each fit is wrapped into a `mdl` — the object that carries a
model's causal context (its formula, its term roles, its degrees of freedom,
its variance–covariance matrix) alongside the fit itself, which is what the
rest of the grammar consumes. `fit(..., raw = TRUE)` instead returns the
plain list of fitted objects the fitting function itself produces — useful
for a quick look, but nothing downstream can use it as a `{epigram}` object.

```{r}
m <- fit(f, .fn = lm, data = d)
m
```

## Collecting models: `mdl_tbl`

A single family already produces several models; a research project produces
many families. `model_table()` (alias `mdl_tbl()`) is the notebook that
stores them:

```{r}
mt <- model_table(m, data = d)
mt
```

The print leads with the fleet's status at a glance — how many fitted, how
many failed, whether data is attached — then one line per model. Passing
`data =` attaches the fitting dataset whole, which the table remembers for
everything downstream that needs it (categorical levels, event counts, the
table layer in the next vignette) — including columns no current formula
names, like a follow-up column or a variable for the next family of models.

`summary()` gives the aggregate view: models grouped by dataset, fitting
function, outcome, and exposure, with their adjustment range and the terms by
causal role.

```{r}
summary(mt)
```

## Stratified fits produce one model per level

A family stratified with `add_strata()` (see `vignette("playing")`) is not
fit once — `fit()` fits one model per stratum level automatically, and the
level lands as provenance in the table:

```{r}
fs <- fmls(mpg ~ .x(wt) + hp) |> add_strata(am) |> set_data(d)
ms <- fit(fs, .fn = lm, data = d)
mts <- model_table(ms, data = d)
mts
```

`mdl_tbl` is a `{dplyr}`-friendly tibble subclass — `filter()`, `select()`,
`arrange()`, and `[` all work, and reconcile the table's internal attributes
(the formula matrix, the term table, the attached data) down to the rows that
remain:

```{r}
dplyr::filter(mts, level == "Manual")
```

## Fitting fails softly

Twenty models in a batch should not sink because one formula references a
column that does not exist. A failure is recorded, not thrown: `fit_status`
is `FALSE`, the error is stored, and the fleet keeps its shape.

```{r}
fbad <- fmls(mpg ~ .x(wt) + not_a_real_column)
mbad <- fit(fbad, .fn = lm, data = d)
mtbad <- model_table(mbad, data = d)
mtbad
model_failures(mtbad)
```

## Mixed models fit through the same interface

A family carrying a `.r()` random-effects term (`vignette("playing")`) fits
directly through a mixed-model engine — no special-casing required:

```{r, eval = requireNamespace("lme4", quietly = TRUE)}
fr <- fmls(mpg ~ .x(wt) + hp + .r(gear))
fit(fr, .fn = lme4::lmer, data = d)
```

## Extracting estimates: `flatten_models()`

`mdl_tbl` stores models; `flatten_models()` unpacks them into one tidy row
per coefficient, ready for analysis or a quick look before building a
publication table:

```{r}
flatten_models(mt)[c("number", "term", "estimate", "conf_low", "conf_high", "p_value")]
```

Exponentiation is inferred, not requested by name: a Cox model or a GLM on a
`log`/`logit`/`cloglog` link comes back as a ratio (hazard, odds, or rate);
everything else stays on the linear scale. `exponentiate = TRUE/FALSE`
overrides the inference when needed.

```{r}
mOr <-
  fmls(am ~ .x(wt)) |>
  fit(.fn = glm, family = stats::binomial, data = d) |>
  model_table(data = d)

flatten_models(mOr)[c("term", "estimate", "exponentiated")]
```

## What's next

A `mdl_tbl` full of fitted models is the raw material for a publication
table. `vignette("mesa")` covers the last layer: laying those models out on
the *mesa* itself, built on `{gt}`.
