---
title: "2. Formulas and Patterns"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Formulas and Patterns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mesa)
```

Layer 2 of the grammar answers the question a single term cannot: how do
terms *combine*, and by what pattern do they expand? A `fmls` ("formulas") is
a **family** — one model idea, expanded into every formula that idea implies.

## From terms to a family

`fmls()` takes the same rune-decorated formula [tm()] does, and expands it by
a **pattern**: the rule for how the covariates enter.

```{r}
f <- fmls(mpg ~ .x(wt) + hp + cyl, pattern = "sequential")
f
```

The deck header up top is a summary — how many formulas, the outcome, the
exposure — the way a `{ggplot2}` plot summarizes its layers. The formulas
themselves come out with [formula()]:

```{r}
formula(f)
```

## The patterns

Assuming `wt`, `hp`, and `cyl` are covariates alongside the exposure `wt`
(here doing double duty as both the exposure and a pattern input), four
built-in patterns answer "how many formulas, and what's in each":

- **`direct`** (the default): every covariate at once, one formula.
- **`sequential`**: covariates enter one at a time, each formula nesting the
  last — the classic "unadjusted, then progressively adjusted" table.
- **`parallel`**: one formula per covariate, each paired with the exposure —
  useful for screening many candidate adjustors independently.
- **`fundamental`**: every right-hand term, the exposure included, entirely
  alone and unadjusted — one formula per term, the univariate screen the
  other patterns build outward from.

```{r}
formula(fmls(mpg ~ .x(wt) + hp + cyl, pattern = "direct"))
formula(fmls(mpg ~ .x(wt) + hp + cyl, pattern = "parallel"))
formula(fmls(mpg ~ .x(wt) + hp + cyl, pattern = "fundamental"))
```

`formula_patterns()` lists every pattern currently available — the four
built-ins plus anything registered:

```{r}
formula_patterns()
```

## Interactions travel with their exposure

A term wrapped `.i()` is a candidate effect modifier. It is crossed with the
exposure automatically, and the pair is grouped so it travels through the
pattern together:

```{r}
fi <- fmls(mpg ~ .x(wt) + .i(am))
formula(fi)
```

## The mediation triad

A term wrapped `.m()` is not just adjusted for — its presence asks a causal
question, so `fmls()` builds the three formulas an epidemiologist needs to
reason about mediation (VanderWeele and Robins 2007), instead of the one
`.x()`/`.c()` would produce:

```{r}
fm <- fmls(mpg ~ .x(wt) + .m(qsec) + hp)
fm
formula(fm)
```

1. `mpg ~ wt + qsec + hp` — the exposure's effect with the pathway through the
   mediator held open (adjusted for the mediator).
2. `qsec ~ wt` — the exposure's effect on the mediator itself.
3. `mpg ~ wt + qsec` — the mediator's effect on the outcome in the exposure's
   presence.

Comparing the exposure's estimate across (1) and (3) is what lets you judge
how much of the total effect travels through the mediator.

## Grouped terms move as one block

Terms that should always enter or leave a formula together — a set of
dummy-coded categories from one variable, say — are tagged with the same
`.g()` tier instead of being split apart by a `sequential` or `parallel`
pattern:

```{r}
fg <- fmls(mpg ~ wt + .g1(hp) + .g1(cyl), pattern = "sequential")
formula(fg)
```

`hp` and `cyl` enter the second formula together, as one adjustment step,
rather than one at a time.

## Combining families

Two families merge with `c()` (or `vctrs::vec_c()`); when the same term
arrives from both sides with conflicting definitions, the left-most
definition wins and a message names the resolution — never a silent
overwrite.

```{r}
f1 <- fmls(mpg ~ .x(wt) + hp)
f2 <- fmls(mpg ~ .x(wt) + cyl)
formula(c(f1, f2))
```

## Patterns are an open registry

The four built-ins are not a closed switch statement — a pattern is any
function from a `tm` vector to a formula-matrix precursor, and
`register_pattern()` adds one by name:

```{r}
unadjusted <- function(x) {
  apply_fundamental_pattern(x)
}
register_pattern("unadjusted", unadjusted)
"unadjusted" %in% formula_patterns()

formula(fmls(mpg ~ .x(wt) + hp + cyl, pattern = "unadjusted"))
```

## What's next

A family of formulas is still just instructions — no data has been
consulted yet. `vignette("playing")` covers how a family learns about a
dataset: strata, subsets, and random effects, the tools that make working
with `{mesa}` feel like *play*.
