---
title: "5. Making the Mesa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5. Making the Mesa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mesa)
```

The destination of the grammar is the fifth layer: laying a collection of
fitted models out for a paper. This is where the package earns its name —
`mdl_gt()` lays the models out on the *mesa*, the table upon which they are
displayed — and it is built as a small composable grammar on top of `{gt}`,
grown the way a `{ggplot2}` plot is grown: one verb, one decision, in any order.

```{r}
d <- mtcars
d$cyl <- factor(d$cyl)

mt <-
  fmls(mpg ~ .x(wt) + hp + cyl, pattern = "sequential") |>
  fit(.fn = lm, data = d) |>
  model_table(data = d)
```

`mt` is the `mdl_tbl` from `vignette("fitting")` — three nested models,
`mpg ~ wt`, `mpg ~ wt + hp`, and `mpg ~ wt + hp + cyl`.

## A bare mesa is already a table

`mdl_gt()` is deliberately bare — it takes no selection arguments — and puts
everything fitted on the mesa with default labels:

```{r}
mt |> mdl_gt()
```

Printing a `<mdl_gt>` specification shows what's on it without rendering
anything: the layout, the current selection, the column blocks so far.
`as_gt()` renders it — and a bare specification already produces a minimal
table, estimate and interval for every displayed term:

```{r}
mt |> mdl_gt() |> as_gt()
```

## Growing the table, one verb at a time

The unit of a mesa is the **term × effect cell**: one term (or term level)
crossed with one effect — an estimate, its interval, a p-value, an event
count, a rate, an observation count. The `add_*()` verbs append *column
groups* of these effects that travel together (an estimate with its CI and
p-value), so groups compose freely: adding, dropping, or reordering one
never changes another's cells, and the layout presets place the same cells
wide (levels on columns) or long (levels on rows) without recomputing
anything.

Nothing has to be decided up front. *Which models* sit on the mesa was
decided before `mdl_gt()`, on the `mdl_tbl` itself — `identify_family()`
stamps the family structure, and the paring verbs speak the causal
language: `keep_outcomes()` / `keep_exposures()` / `drop_*()`,
`keep_families()`, `restrict_to()`, `adjusting_for()`, `excluding()`
(ordinary `dplyr` verbs work too). The specification's verbs make *display*
decisions, each narrowing or adding exactly one thing, composing in any
order:

- **`select_adjustment()` / `select_terms()`** choose which adjustment rungs
  and which terms' cells are displayed (`adjustment_sets()` shows the rungs
  and their numbers), and their input is a labeled formula — so selecting
  and labeling are one gesture.
- **`modify_labels()`** relabels outcomes, terms, levels, or columns late,
  without reselecting anything.
- **`add_estimates()`, `add_n()`, `add_events()`, `add_rate_difference()`,
  `add_forest()`, `add_interaction()`** each append one column block.
- **`modify_layout()`, `modify_style()`** assign the table's axes and its
  accents, digits, and padding.

```{r}
mt |>
  mdl_gt() |>
  select_adjustment(1 ~ "Unadjusted", 3 ~ "Fully adjusted") |>
  add_estimates(columns = list(beta ~ "Estimate", conf ~ "95% CI")) |>
  modify_labels(wt ~ "Weight (1000 lbs)") |>
  as_gt()
```

Because a `<mdl_gt>` specification is *declarative* — verbs record instructions,
and resolution happens only at `as_gt()` — repeating a verb replaces just that
instruction, with a message, rather than compounding silently. Rethinking a
label, or narrowing the selection further, is always cheap.

## The three layout presets

A layout **preset** is a complete assignment of the grammar's four axes
(rows, row groups, columns, spanners), selected with `modify_layout()`.

### `"adjustment"` (the default)

Adjustment sets on rows, outcomes as row groups — the shape above, and the
one every bare mesa starts from.

### `"levels"`

Event counts and incidence rates, then adjusted-estimate rows, with term
levels on columns — the shape of a hazard-ratio table:

```{r, eval = requireNamespace("survival", quietly = TRUE)}
lung <- survival::lung
lung$sex <- factor(lung$sex, levels = 1:2, labels = c("Male", "Female"))

mtSurv <-
  fmls(Surv(time, status) ~ .x(sex) + age, pattern = "sequential") |>
  fit(.fn = survival::coxph, data = lung) |>
  model_table(data = lung)

mtSurv |>
  mdl_gt() |>
  modify_layout(preset = "levels") |>
  select_adjustment(1 ~ "Unadjusted", 2 ~ "Age-adjusted") |>
  add_events(followup = time) |>
  add_rate_difference() |>
  add_estimates(columns = list(beta ~ "HR", conf ~ "95% CI")) |>
  as_gt()
```

`add_events()` reads the follow-up column from the `Surv()` outcome itself
when it can (`followup = time` above makes it explicit; it is optional when
the data's follow-up column matches the outcome's time argument). Because a
Cox model's estimates come back exponentiated by default (the family
inference from `vignette("fitting")`), the column is already labeled `HR` on
the correct scale.

### `"interaction"`

Interaction levels on rows, grouped by interaction term, with the
across-levels p-value floating over each band — effect modification made
visible:

```{r}
mtInt <-
  fmls(mpg ~ .x(hp) + .i(cyl)) |>
  fit(.fn = lm, data = d) |>
  model_table(data = d)

mtInt |>
  mdl_gt() |>
  add_interaction() |>
  add_n(label = "No.") |>
  add_estimates(columns = list(beta ~ "Estimate", conf ~ "95% CI",
                                p ~ "P for interaction")) |>
  as_gt()
```

`add_interaction()` **implies** the `"interaction"` layout — it defines the
rows, so declaring the block is declaring the layout; a separate
`modify_layout(preset = "interaction")` call is never required for the common
case.

## A forest column, on any table

`add_forest()` appends a forest-plot column to any table that already carries
an estimate and interval — it reads them and computes nothing new, so adding
or dropping it never changes any other cell:

```{r}
mtInt |>
  mdl_gt() |>
  add_interaction() |>
  add_estimates() |>
  add_forest() |>
  as_gt()
```

## Style: accents, digits, missing text

`modify_style()` emphasizes cells that meet a criterion — any displayed
statistic, not just a p-value — and controls formatting table-wide:

```{r}
mt |>
  mdl_gt() |>
  add_estimates() |>
  modify_style(accents = list(p < 0.05 ~ "bold"), digits = 3) |>
  as_gt()
```

## What's next

That is the whole grammar: `mdl_gt()` to start, `select_*()` to narrow,
`add_*()` to add statistics, `modify_*()` to lay out and style, `as_gt()` to
render. To step back from the mechanics to the reasoning the roles are for —
how variable roles map onto estimands like the total effect, direct effect,
and effect modification — start from `vignette("terms", package = "mesa")`,
where the role vocabulary is laid out.
