---
title: "5. Making the Mesa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5. Making the Mesa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mesa)
```

The destination of the grammar is the fifth layer: laying a collection of
fitted models out for a paper. This is the package's namesake act — `mesa()`
lays models on the *mesa*, the table upon which they are displayed — and it
is built as a small composable grammar on top of `{gt}`, grown the way a
`{ggplot2}` plot is grown: one verb, one decision, in any order.

```{r}
d <- mtcars
d$cyl <- factor(d$cyl)

mt <-
  fmls(mpg ~ .x(wt) + hp + cyl, pattern = "sequential") |>
  fit(.fn = lm, data = d) |>
  model_table(data = d)
```

`mt` is the `mdl_tbl` from `vignette("fitting")` — three nested models,
`mpg ~ wt`, `mpg ~ wt + hp`, and `mpg ~ wt + hp + cyl`.

## A bare mesa is already a table

`mesa()` is deliberately bare — it takes no selection arguments — and puts
everything fitted on the mesa with default labels:

```{r}
mt |> mesa()
```

Printing a `<mesa>` specification shows what's on it without rendering
anything: the layout, the current selection, the column blocks so far.
`as_gt()` renders it — and a bare specification already produces a minimal
table, estimate and interval for every displayed term:

```{r}
mt |> mesa() |> as_gt()
```

## Growing the table, one verb at a time

Nothing has to be decided up front. Each verb narrows or adds exactly one
thing, and verbs compose in any order:

- **`select_outcomes()` / `select_exposures()` / `select_terms()` /
  `select_adjustment()` / `select_strata()`** narrow what is shown, and their
  input is a labeled formula — so selecting and labeling are one gesture.
- **`modify_labels()`** relabels terms, levels, or columns late, without
  reselecting anything.
- **`add_estimates()`, `add_n()`, `add_events()`, `add_rate_difference()`,
  `add_forest()`, `add_interaction()`** each append one column block.
- **`modify_layout()`, `modify_style()`** assign the table's axes and its
  accents, digits, and padding.

```{r}
mt |>
  mesa() |>
  select_adjustment(1 ~ "Unadjusted", 3 ~ "Fully adjusted") |>
  add_estimates(columns = list(beta ~ "Estimate", conf ~ "95% CI")) |>
  modify_labels(wt ~ "Weight (1000 lbs)") |>
  as_gt()
```

Because a `<mesa>` specification is *declarative* — verbs record instructions,
and resolution happens only at `as_gt()` — repeating a verb replaces just that
instruction, with a message, rather than compounding silently. Rethinking a
label, or narrowing the selection further, is always cheap.

## The three layout presets

A layout **preset** is a complete assignment of the grammar's four axes
(rows, row groups, columns, spanners), selected with `modify_layout()`.

### `"adjustment"` (the default)

Adjustment sets on rows, outcomes as row groups — the shape above, and the
one every bare mesa starts from.

### `"levels"`

Event counts and incidence rates, then adjusted-estimate rows, with term
levels on columns — the shape of a hazard-ratio table:

```{r, eval = requireNamespace("survival", quietly = TRUE)}
lung <- survival::lung
lung$sex <- factor(lung$sex, levels = 1:2, labels = c("Male", "Female"))

mtSurv <-
  fmls(Surv(time, status) ~ .x(sex) + age, pattern = "sequential") |>
  fit(.fn = survival::coxph, data = lung) |>
  model_table(data = lung)

mtSurv |>
  mesa() |>
  modify_layout(preset = "levels") |>
  select_adjustment(1 ~ "Unadjusted", 2 ~ "Age-adjusted") |>
  add_events(followup = time) |>
  add_rate_difference() |>
  add_estimates(columns = list(beta ~ "HR", conf ~ "95% CI")) |>
  as_gt()
```

`add_events()` reads the follow-up column from the `Surv()` outcome itself
when it can (`followup = time` above makes it explicit; it is optional when
the data's follow-up column matches the outcome's time argument). Because a
Cox model's estimates come back exponentiated by default (the family
inference from `vignette("fitting")`), the column is already labeled `HR` on
the correct scale.

### `"interaction"`

Interaction levels on rows, grouped by interaction term, with the
across-levels p-value floating over each band — effect modification made
visible:

```{r}
mtInt <-
  fmls(mpg ~ .x(hp) + .i(cyl)) |>
  fit(.fn = lm, data = d) |>
  model_table(data = d)

mtInt |>
  mesa() |>
  add_interaction() |>
  add_n(label = "No.") |>
  add_estimates(columns = list(beta ~ "Estimate", conf ~ "95% CI",
                                p ~ "P for interaction")) |>
  as_gt()
```

`add_interaction()` **implies** the `"interaction"` layout — it defines the
rows, so declaring the block is declaring the layout; a separate
`modify_layout(preset = "interaction")` call is never required for the common
case.

## A forest column, on any table

`add_forest()` appends a forest-plot column to any table that already carries
an estimate and interval — it reads them and computes nothing new, so adding
or dropping it never changes any other cell:

```{r}
mtInt |>
  mesa() |>
  add_interaction() |>
  add_estimates() |>
  add_forest() |>
  as_gt()
```

## Style: accents, digits, missing text

`modify_style()` emphasizes cells that meet a criterion — any displayed
statistic, not just a p-value — and controls formatting table-wide:

```{r}
mt |>
  mesa() |>
  add_estimates() |>
  modify_style(accents = list(p < 0.05 ~ "bold"), digits = 3) |>
  as_gt()
```

## What's next

That is the whole grammar: `mesa()` to start, `select_*()` to narrow,
`add_*()` to add statistics, `modify_*()` to lay out and style, `as_gt()` to
render. To step back from the mechanics to the reasoning the roles are for —
how variable roles map onto estimands like the total effect, direct effect,
and effect modification — start from `vignette("terms", package = "mesa")`,
where the role vocabulary is laid out.
