---
title: "3. Playing with Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Playing with Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(epigram)
```

The feeling `{epigram}` is meant to give is *fluidity* — picking terms up off a
dataset and playing with the modeling: swap strata, add a random effect,
subset the data, roll an interaction through, and watch the family of models
unfold. This vignette covers the fluent verb layer that makes that possible,
and [set_data()], the step where a family finally meets a dataset.

```{r}
d <- mtcars
d$cyl <- factor(d$cyl)
d$am <- factor(d$am, labels = c("Automatic", "Manual"))
```

## The fluent verbs

Every verb here is pipeable, and every one returns a modified `fmls` — so
they compose into a single readable chain, the way `{dplyr}` verbs do.

```{r}
f <- fmls(mpg ~ .x(wt) + hp)

f |>
  add_strata(am) |>
  add_terms(cyl) |>
  subset_data(disp > 100)
```

- **`add_strata()` / `remove_strata()`** promote a term to (or release it
  from) the `strata` role. A stratified family is not fit once — `fit()`
  fits one model per stratum level, automatically.
- **`add_terms()` / `remove_terms()`** add or drop covariates from every
  formula in the family at once.
- **`swap_outcome()`** exchanges one outcome for another, keeping the rest of
  the family — its exposure, its covariates, its pattern — intact. Either
  `old ~ new`, or a bare replacement name when the family has a single
  outcome.
- **`subset_data()`** records data-filtering instructions — plain logical
  expressions in the data's variables — that `fit()` applies later, fitting
  the family once per subset. The filter is not applied here; it is *recorded*
  as an instruction, the same way a stratum level is.

```{r}
f |> swap_outcome(qsec)
```

## Meeting the data

Everything above is structural — roles, strata, subsets — decided before any
dataset is consulted. [set_data()] is the step where a family actually meets
one: it classifies every term's `type` and `distribution`, and — this is the
detail that matters for strata — stamps the *observed levels* onto any
categorical term, strata included.

Call it **last**, after the roles it needs to classify are already assigned:

```{r}
f2 <-
  fmls(mpg ~ .x(wt) + hp) |>
  add_strata(am) |>
  set_data(d)

f2
```

The deck header now reads `strata: am (2 levels)` — before a single model has
been fit, the family already knows how many models a stratified fit will
produce. Calling `set_data()` *before* `add_strata()` would classify `am` as
a plain predictor first, so the strata levels would not appear; `set_data()`
reflects the roles that exist at the moment it runs, which is why it reads
naturally as the last step in a chain of structural verbs.

## Subsets are not a role

A filter like `am == "Manual"` is not a variable in the formula — it is an
instruction about the data, so `subset_data()` deliberately does not reuse
any of the causal roles. It records the instruction on the family, prints in
the deck header, and expands the fitting plan exactly like a stratum level
does: one model per subset (and, combined with strata, one per
stratum-by-subset combination).

```{r}
f3 <-
  fmls(mpg ~ .x(wt) + hp) |>
  subset_data(am == "Manual", cyl == 4) |>
  set_data(d)

f3
```

## Random effects

Random (hierarchical) grouping is a role like any other — `.r()` — and it
composes, prints, and rebuilds into the `(1 | term)` syntax mixed-model
engines expect. `{epigram}` also parses that native `lme4` syntax directly, so
both spellings land on the same term:

```{r}
fr <- fmls(mpg ~ .x(wt) + hp + .r(gear))
formula(fr)

fr_native <- fmls(mpg ~ .x(wt) + hp + (1 | gear))
identical(formula(fr), formula(fr_native))
```

Unlike a plain covariate, a random-effects term is excluded from covariate
*expansion* — a `sequential` or `parallel` pattern never rotates it in or out
the way it does `hp` or `cyl` — but it rides along on every formula the
pattern produces.

## What's next

A family that knows its strata, its subsets, and the shape of its data is
ready to actually meet a fitting function. `vignette("fitting")` covers
`fit()`, the plan it builds, and the `mdl_tbl` collection that stores the
resulting models.
