---
title: "1. Terms and Causal Roles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Terms and Causal Roles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mesa)
```

`{mesa}` is built from an epistemological perspective: small building blocks
that build up to more complex concepts.
The smallest block is the **term** — a single variable, carrying not just its
name but its causal identity.
This vignette covers layer 1 of the grammar: what a term is, the roles it can
play, and how it learns about the data it will eventually be fit against.

## A term is more than a name

A plain `R` formula only knows a variable's *position* — left of the tilde or
right of it.
It has no way to say that `wt` is the exposure whose effect you care about,
that `cyl` is a confounder you are adjusting for, or that `am` defines
subpopulations you want fit separately.
The `tm` class fills that gap.

```{r}
t <- tm(mpg ~ wt + hp)
t
```

Printed, a `tm` vector looks almost like the formula it came from — but every
element already carries a role (here, everything but `mpg` defaulted to
`predictor`, and `mpg` was inferred as the `outcome` from its position on the
left-hand side).

## The role taxonomy

Roles are causal, and they are first-class: a term's role is knowable,
updatable, and visibly changes what happens downstream — how formulas expand
([fmls()]), how models are fit ([fit()]), and how results are labeled once
they reach a table ([mesa()]).

```{r}
term_roles()
```

| Role | Shortcut | Definition | Downstream behavior |
| --- | --- | --- | --- |
| outcome | `.o(...)` | the dependent variable; the effect being studied | anchors the left-hand side; multiple outcomes multiply the formula family |
| exposure | `.x(...)` | the variable whose causal effect is under study | anchored in every expanded formula; pairs with interactions |
| predictor | `.p(...)` | a covariate with no asserted causal position | expanded by the chosen pattern (adjusted for, or rotated through) |
| confounder | `.c(...)` | a common cause of exposure and outcome | treated as a covariate; flagged for adjustment displays |
| mediator | `.m(...)` | on the causal pathway between exposure and outcome | triggers the mediation triad of formulas |
| interaction | `.i(...)` | a candidate effect modifier of the exposure | crossed with each exposure (`x:i`), grouped so the pair travels together |
| strata | `.s(...)` | a variable defining subpopulations for separate fits | not a covariate; `fit()` fits one model per stratum level |
| random | `.r(...)` | a grouping variable for random (hierarchical) effects | rendered as `(1 \| term)` for mixed-model engines |
| group | `.g(...)` | not a role, but a tier marker for terms that travel together | grouped terms enter and leave expanded formulas as one block |

A role is applied by wrapping the variable in its shortcut, right inside the
formula:

```{r}
t <- tm(mpg ~ .x(wt) + hp + .c(cyl))
t
describe(t, "role")
```

`hp` was left bare and defaulted to `predictor` — the role assumed for any
right-hand term with no asserted causal position, and the one a pattern
rotates through or adjusts for. Nothing about this is permanent: roles are
updated after the fact with [update.tm()], just like a label is.

```{r}
t2 <- update(t, role = hp ~ "confounder")
describe(t2, "role")
```

## Transformations round-trip losslessly

Wrapping a term in a recognized transformation — `log()`, `sqrt()`, `scale()`,
`factor()`, `ordered()`, polynomial and spline markers — keeps the *full call*
as the term's name, so a formula rebuilds exactly, without `{mesa}` needing to
re-apply anything itself. The wrapper is additionally recorded for downstream
interpretation.

```{r}
t3 <- tm(mpg ~ .x(log(wt)) + hp)
t3
describe(t3, "transformation")
formula(t3)
```

`term_transformations()` lists every wrapper `{mesa}` recognizes; an
unrecognized call (`survival::Surv(time, status)`, say) is carried whole as an
opaque term rather than rejected.

## Meeting the data

A term's role is knowable before it ever sees data, but its *shape* —
continuous or categorical, and which levels a categorical term actually
has — is not. [set_data()] is the "meet the data" step: it classifies every
term against a data frame and stamps `type`, `distribution`, and `level`
directly onto it.

```{r}
d <- mtcars
d$cyl <- factor(d$cyl)

t4 <- tm(mpg ~ .x(wt) + hp + .c(cyl)) |> set_data(d)

# vctrs::vec_proxy() exposes a term vector's full attribute table
vctrs::vec_proxy(t4)[c("term", "role", "type", "level")]
```

`cyl` is now known to be categorical with three observed levels — information
the table layer ([mesa()]) later needs to decide how many columns a
categorical term needs and which level is the reference. A transformed term
classifies from its *underlying* variable: `set_data()` looks through
`log(wt)` to `wt` itself when deciding whether it is continuous.

## Labeling

A term's `label` is what a table eventually prints instead of its bare
variable name — set at creation, or rethought later with the same
[update.tm()] verb used for roles:

```{r}
t5 <- update(t4, label = wt ~ "Weight (1000 lbs)")
describe(t5, "label")
```

## What's next

A single term is an atom; the next vignette, `vignette("formulas")`, covers
how terms join into **families of formulas** — one model idea, expanded by a
pattern into the several formulas that idea implies.
