---
title: "Internals: how the machine is built"
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Developer-facing companion to `vignette("usage")`. Shows the data structure
under each layer and the seams between them, using `mesa:::` internals where
needed. Each section ends with **review questions** — the places where the
design deserves a second pair of eyes. Concrete defects live in `issues.md`.

```{r setup}
library(mesa)
d <- mtcars
d$cyl <- factor(d$cyl)
```

## 1. A term is a record

`tm` is a `vctrs` record: parallel fields, one slot per term. The proxy *is*
the term table — every later structure is built from this.

```{r}
t <- tm(mpg ~ .x(wt) + .s(am) + log(hp))
vctrs::vec_proxy(t)
```

`tm.formula()` is a fixed pipeline over the syntax tree:

1. `parse_formula_terms()` — recursive descent; runes, `a:b`, `(1|g)`,
   transformations, opaque calls (`Surv(...)`) each handled at the node
2. `demote_orphan_roles()` — `.i()`/`.m()` without an exposure demote to
   predictors, with a warning
3. `expand_shortcut_interactions()` — each `.i()` crosses each exposure into
   an explicit `x:i` term, group-tiered so the pair travels together
4. `apply_default_roles()` — position fills in what wasn't declared
5. `apply_term_arguments()` — explicit `role = `/`label = ` arguments override

The raw parse, before defaults:

```{r}
mesa:::parse_formula_terms(mpg ~ .x(wt) + .i(cyl))
```

Note `side`: strata and random effects get `side = "meta"` — neither left nor
right. That single value drives everything in §3.

**Review questions**

- `log(hp)` keeps its full call as its *name* and records the wrapper in
  `transformation`. Two representations of one fact — is the field earning
  its keep, or should the call be parsed once at `set_data()`?
- Roles live in a flat character field; `group` is a separate integer tier.
  Interactions need both to travel together. Simpler as one composite?

## 2. A `fmls` is a membership matrix plus a term table

Two structures, stitched by term name:

```{r}
f <- fmls(mpg ~ .x(wt) + hp, pattern = "sequential")
vctrs::vec_data(f)      # rows = formulas, cols = terms, 1 = membership
attr(f, "termTable")    # scalar attribute: one definition per term
```

Patterns are functions `tm -> tbl_df` (a registry, `register_pattern()`);
their output is a *precursor table*, one row per future formula:

```{r}
apply_sequential_pattern(tm(mpg ~ .x(wt) + hp))
```

`fmls()` then flattens each precursor row into the 0/1 matrix directly —
each row's unique non-`NA` terms become that row's memberships, so a term
repeated within one precursor row still records `1`.

Every registered pattern honors that contract now,
`apply_fundamental_pattern()` included: the exposure keeps its key-pair
column and each remaining term rides one row as `covariate_1`. Because the
fundamental pattern reduces each formula to a single left- and right-hand
term, `fmls(pattern = "fundamental")` demotes meta terms (strata, random
effects) to plain predictors before expansion, with a message:

```{r}
fmls(mpg ~ wt + .s(am), pattern = "fundamental")
```

**Review questions** *(both resolved 2026-07-09)*

- ~~The seven-role extraction block is copy-pasted through every pattern.~~
  `pattern_roles()` pulls every role once and `key_pair_grid()` builds the
  outcome × exposure grid; the patterns, `check_mediation()`, and
  `check_groups()` all draw from them.
- ~~The `table()`-based matrix build records a count of 2 for a repeated
  term.~~ The membership build above is 0/1 by construction, and the
  downstream tests read `>= 1`.

## 3. Where the strata goes (and why formulas drop it)

`formula.tm()` rebuilds from `side == "left"` and `side == "right"` only.
Random effects are re-added as `(1 | g)`; **strata is not re-added anywhere**:

```{r}
f <- fmls(mpg ~ .x(wt) + .s(am))
formula(f)
```

The stratum re-enters at `fit_plan()` — not as formula syntax but as *data
splitting*: one plan row per stratum level, the level's rows subset before
the engine sees anything.

```{r}
fit_plan(f, data = d)
```

Meta terms are members of the formula matrix like any other term, and
`formulas_to_terms()` reads membership alone — so a stratum applies only to
the family that declared it, and combining families cannot leak it:

```{r}
combined <- suppressMessages(c(f, fmls(qsec ~ .x(hp))))
fit_plan(combined, data = d)[, c("formula_call", "strata_variable")]
```

**Review questions**

- ~~The stratum is invisible everywhere models print.~~ *Resolved
  2026-07-09*: the printed family annotates meta terms in rune form
  (`.s(am)`, `.r(id)`) after the right-hand terms, while `formula()`,
  `formula_call`, and the recorded call stay the true fitted formula —
  they are consumed programmatically (re-fitting, term counting), so
  annotating them would corrupt real syntax. The `strata`/`level` columns
  remain the machine-readable context.
- Some engines have native stratification (`survival::strata()` inside a
  `coxph` formula — a *conditioning* term, not a data split). The two
  meanings share a word but not a mechanism. Do we need both, and how do
  they coexist at the fitting stage? *(Still open; tracked in `issues.md`.)*

## 4. Why data attaches to the *table*, not the formula

Formulas stay abstract on purpose — a `fmls` captures no data, so families
are cheap to build, combine, and store. But three consumers need real data
at realization time:

- **categorical levels** — the term table's `level` field is empty until
  `set_data()`; the table layer stamps it from attached data to know that
  `cyl8` is level `8` of `cyl` (`resolve_term_metadata()`)
- **events and rates** — `add_events()` runs `survival::pyears()` on the
  attached data at render
- **interaction rows** — per-level `n` counts come from the attached data

So the dataset attaches once, to the `mdl_tbl`, keyed by name:

```{r}
mt <-
  fmls(mpg ~ .x(wt) + hp + cyl, pattern = "sequential") |>
  fit(.fn = lm, data = d) |>
  model_table(data = d)

names(attr(mt, "dataList"))
term_table(mt)
```

**Review questions**

- ~~The link is a fragile *name* match.~~ *Resolved 2026-07-09*: a `data`
  passed as an inline expression takes a stable content-derived id
  (`data_<hash>`) at both `fit()` and `attach_data()`, so identical content
  meets itself; and a frame arriving under a different name is aliased —
  with a message — to the one referenced-but-detached `data_id` whose
  models' variables it fully carries (`data_id_candidates()`). An explicit
  `name = ` always wins.
- Attaching whole data frames makes a `mdl_tbl` heavy to serialize. Store
  only the columns the terms reference?

## 5. A `mdl` is a deconstructed model

Not the fitted object — a blueprint: call name, `fmls`, tidy estimates,
glance summary, data context. Enough to present and diagnose, cheap to keep.

```{r}
m <- fit(fmls(mpg ~ .x(wt) + hp), .fn = lm, data = d)
vctrs::field(m, "parameterEstimates")[[1]]
vctrs::field(m, "dataArgs")[[1]]
```

Notes:

- `execute_fit()` rewrites the recorded call (function by name, data by
  name) so models hash and compare cleanly.
- Failures are soft: an error becomes a `mdl` whose `summary_info$error`
  carries the message; `fit_status` turns `FALSE` downstream.
- The `var_cov` matrix is stored per model — that is what lets
  `estimate_interaction()` derive within-level effects without refitting.

**Review questions** *(both resolved 2026-07-09)*

- ~~`mdl.lm` and `mdl.lmerMod` are ~80% identical.~~ Both are thin
  wrappers now: each computes its own tidy/glance pieces and hands the
  rest to `new_model_from_fit()`, the shared trunk.
- ~~`my_tidy()` exposes an `exponentiate` argument it ignores.~~ The
  parameter is gone; exponentiation stays deferred to `flatten_models()`.

## 6. The `mdl_tbl`: invariant columns, reconciled attributes

A tibble subclass — one row per model — carrying three scalar attributes
that must stay parallel to the rows: the formula matrix, the term table, the
data list. Every subset/filter reconciles them (`df_reconstruct()`):

```{r}
sub <- dplyr::filter(mt, number < 3)
class(sub)
formula_matrix(sub)      # trimmed to the surviving rows and their terms
```

Dropping an invariant column deliberately demotes the object:

```{r}
class(dplyr::select(mt, -fit_status))
```

**Review questions**

- ~~`formula_index` stores the membership row itself, and nothing consumes
  it.~~ *Resolved 2026-07-09*: dropped. The formula matrix's rows stay
  parallel to the table's, so a row-index column carried no information.
- Rows link to attributes by `id = rlang::hash(model)`. Two identical fits
  (same formula, same data) collide by design — is dedup ever wrong?

## 7. The mesa: five stages from spec to `{gt}`

A `<mesa>` is a plain list of *instructions* — nothing is computed until
render:

```{r}
spec <- mt |> mesa() |> add_estimates()
spec$selection
spec$columns
spec$layout
```

Realization runs in fixed stages, each file owning one:

| stage | function | file |
|---|---|---|
| select | `resolve_selection()` | table-selection.R |
| decorate | `realize_mesa()` | table-realize.R |
| compute | `compute_data_statistics()` | table-columns.R |
| layout | `mesa_frame()` → cell frame | table-presets.R |
| render | `render_cell_frame()` | table-render.R |

The decorated frame — one row per displayed estimate, reference rows
injected for categorical terms:

```{r}
mesa:::realize_mesa(spec)[, c("outcome", "adj_index", "variable", "level",
                              "estimate", "is_reference", "adj_label")]
```

The **cell frame** — the single contract between layout and render, one row
per rendered cell; `render_cell_frame()` is the only place `{gt}` calls are
emitted:

```{r}
mesa:::mesa_frame(spec)
```

Two mechanisms live only in the renderer:

- **Group-scoped cells** (the interaction p-value floating over its level
  rows): `{gt}` has no body rowspan, so the value is written into every row
  and all but one copy is *blanked by a text transform*
  (`apply_group_scoped()`) — a content substitution, which holds on dark
  themes and every output format (the old white, zero-size text styling
  did not).
- **Plot cells** (the forest column): each cell is a ggplot drawn as vector
  SVG and inlined as its own base64 `data:` document, sized in `em` units
  so it scales with the text beside it (`plot_image()`; a build without
  cairo falls back to the fixed-size PNG).

**Review questions**

- The adjustment preset emits a `::p` column per term *level*; wide tables
  get very wide. Is a per-term p-column the better default?
- ~~`frame_context()` is called with `dec = NULL` on the interaction
  path.~~ *Resolved 2026-07-09*: the unused parameter is gone;
  `frame_context(spec)` takes the specification alone.

## 8. Review agenda

Ordered start-points, cross-referenced to `issues.md`. The 2026-07-09
repair pass closed the outstanding items: the high-severity contract fixes
(the fundamental-pattern contract, the strata leak on family combination,
the two definitions of "categorical", the interaction layout and estimator),
the medium design-friction items (`fit(raw = )` now defaults to `FALSE`,
strata annotate in printed formulas, forest cells draw as scalable SVG, the
group mask survives themes and formats, the one-family check compares
links, data attaches by content when names diverge), and the parsimony
passes (§2, §5, the sequential pattern, `rhs()`'s tree split).

Still open, by choice:

1. Engine-native strata (`survival::strata()` in a `coxph` formula) versus
   `.s()` data splitting — a design question, not a defect
2. Per-term versus per-level `::p` columns in the adjustment preset
3. Serializing attached data frames (store only referenced columns?)
