---
title: "Usage: terms → formulas → models → tables"
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Working surface for interactive development. One continuous walkthrough of the
pipeline, layer by layer — point at a chunk, say what looks wrong, we repair
it. Succinct by intent; the machinery behind each chunk is in
`vignette("internals")`.

```{r setup}
library(mesa)
d <- mtcars
d$cyl <- factor(d$cyl)
d$am <- factor(d$am, levels = c(0, 1), labels = c("Automatic", "Manual"))
```

## 1. Terms

Roles declared inline with runes (`.x()` exposure, `.o()` outcome, `.s()`
strata, `.i()` interaction, `.m()` mediator, ...):

```{r}
t <- tm(mpg ~ .x(wt) + hp + .s(am))
t
describe(t, "role")
```

Terms become data-aware when they meet a dataset — type, distribution, and
levels get stamped on:

```{r}
t <- set_data(t, d)
describe(t, "level")
```

## 2. Formulas

A `fmls` is a *family*: one recipe, many formulas, expanded by a pattern.

```{r}
fmls(mpg ~ .x(wt) + hp + cyl, pattern = "sequential")
```

```{r}
fmls(mpg ~ .x(wt) + hp + cyl, pattern = "parallel")
```

The listing *is* the display: roles show through the term coloring rather
than a legend, and `describe()` reads them off a family directly:

```{r}
describe(fmls(mpg ~ .x(wt) + hp), "role")
```

Strata ride along as *meta* terms — annotated in the printed family with
their declared rune, absent from the fitted formula itself:

```{r}
f <- fmls(mpg ~ .x(wt) + hp + .s(am))
f
formula(f)
```

> ✓ **Resolved (2026-07-09).** The stratum splits data at `fit()` — it is not
> a covariate — so `formula()` and the `formula_call` column stay the true
> fitted (unstratified) formula. The *printed* family now carries `.s(am)`
> after the right-hand terms (random effects annotate as `.r(...)` the same
> way), so a stratified model can no longer be misread as unstratified.

Interactions cross with the exposure automatically:

```{r}
fmls(mpg ~ .x(hp) + .i(cyl))
```

A mediator triggers the causal triad:

```{r}
fmls(mpg ~ .x(wt) + .m(hp) + qsec)
```

Families are playable after the fact:

```{r}
fmls(mpg ~ .x(wt) + hp) |>
  add_strata(am) |>
  subset_data(disp > 100)
```

## 3. Models

The plan comes before the fit — formula × stratum × subset, inspectable:

```{r}
fs <- fmls(mpg ~ .x(wt) + hp, pattern = "sequential") |> add_strata(am)
fit_plan(fs, data = d)
```

`fit()` keeps the causal context attached by default (a `mdl` vector;
`raw = TRUE` opts out for a quick look at the plain fits); `model_table()`
is the notebook of models:

```{r}
m <- fit(fs, .fn = lm, data = d)
mt <- model_table(m, data = d)
mt
```

> ✓ **Resolved (2026-07-09).** `raw = FALSE` is now the default, so the main
> path no longer repeats the argument at every call.

```{r}
summary(mt)
```

Estimates, one row per parameter:

```{r}
flatten_models(mt)
```

## 4. Tables — the mesa

```{r}
mt2 <-
  fmls(mpg ~ .x(wt) + hp + cyl, pattern = "sequential") |>
  fit(.fn = lm, data = d) |>
  model_table(data = d)
```

A bare `mesa()` is already a table; printing shows the specification without
rendering:

```{r}
mt2 |> mesa()
mt2 |> mesa() |> as_gt()
```

Grown one verb at a time — select, add, modify, in any order:

```{r}
mt2 |>
  mesa() |>
  select_adjustment(1 ~ "Unadjusted", 3 ~ "Fully adjusted") |>
  add_estimates(columns = list(beta ~ "Estimate", conf ~ "95% CI")) |>
  modify_labels(wt ~ "Weight (1000 lbs)") |>
  as_gt()
```

### The `"levels"` layout: events, rates, hazard ratios

```{r}
#| eval: !expr requireNamespace("survival", quietly = TRUE)
lung <- survival::lung
lung$sex <- factor(lung$sex, levels = 1:2, labels = c("Male", "Female"))

mtSurv <-
  fmls(Surv(time, status) ~ .x(sex) + age, pattern = "sequential") |>
  fit(.fn = survival::coxph, data = lung) |>
  model_table(data = lung)

mtSurv |>
  mesa() |>
  modify_layout(preset = "levels") |>
  select_adjustment(1 ~ "Unadjusted", 2 ~ "Age-adjusted") |>
  add_events(followup = time) |>
  add_rate_difference() |>
  add_estimates(columns = list(beta ~ "HR", conf ~ "95% CI")) |>
  as_gt()
```

### The `"interaction"` layout, with a forest column

```{r}
mtInt <-
  fmls(mpg ~ .x(hp) + .i(cyl)) |>
  fit(.fn = lm, data = d) |>
  model_table(data = d)

mtInt |>
  mesa() |>
  add_interaction() |>
  add_n(label = "No.") |>
  add_estimates(columns = list(beta ~ "Estimate", conf ~ "95% CI",
                               p ~ "P for interaction")) |>
  add_forest() |>
  as_gt()
```

> ✓ **Resolved (2026-07-09).** Forest cells now draw as inline vector SVG
> sized in `em` units, so they scale with the text they sit beside (a build
> without cairo falls back to the old fixed-size PNG). See
> `vignette("internals")` §7.

### Style

```{r}
mt2 |>
  mesa() |>
  add_estimates() |>
  modify_style(accents = list(p < 0.05 ~ "bold"), digits = 3) |>
  as_gt()
```
